Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-07-19 paper/paper.tex
Friday WIP
WIP, towards first results figure

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum.
%
\citet{simpson_simple_2011} model three contributions to the movement of the growth
cones; chemoaffinity, competition for space and axon-axon interactions. The
first two of these are modelled without reference to a particular mechanism;
each agent is provided with globally-supervised information about its
instantaneous location and its target location, allowing it to move along a
vector towards the target. Competition is implemented as a movement of
constant distance along the vector between two agents if those two agents are
within a threshold distance of one another. Interaction is signal-induced
repulsion, which is triggered if two agents are both within the threshold
distance of each other and have compatible EphA receptor expression.

In our model, space-based competition is omitted and chemoaffinity and
axon-axon interactions occur only via signal transmission between receptors
and ligands, which are expressed in graded patterns on both retina and
tectum. Signals induce attractive or repulsive interactions; for chemoaffinity
the interaction is to ascend or descend the gradient of ligand expression on
the tectum; for axon-axon interactions, the strength of short-range repulsion
(or attraction) is governed by the receptor-ligand expression on pairs of axons.

\subsection{Receptor and ligand expression}

@@

Each retinal ganglion cell projects $n=8$ growth cone agents (referred to
as \emph{branches}) which carry a set of 4 receptors, $r_i$, and 4 ligands, $l_i$,
expressed at levels determined by the cell's location on the retina. The
expression levels of the receptors and ligands vary with respect to the cell's
position on the retinal surface. Each receptor or ligand varies with respect
only to one dimension.  Receptor expression gradients are arranged in
orthogonal pairs with the gradient of $r_0$ being orthogonal to that of
$r_1$. $r_2$, whose gradient is opposite to $r_0$, is orthogonal to
$r_3$. Ligands are also arranged in orthogonal pairs, with $l_0$ opposing
$r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$ orthogonal to $l_0$ and
opposing $r_1$.

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
The simulated tectum expresses the same types of receptors and ligands, also
in orthogonal pairs of gradients. In this work, we did not consider the effect
of reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression, $L$. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
\textbf{X}). Correspondingly, we leave open the possibility that tectal ligand
expression may be modelled by exponential, linear or logarithmic functions.

@@

\subsection{Chemotaxis}

@@

\subsection{Axon-axon interactions}

For axon-axon interactions, repulsion causes a movement of branch $b$ along a
unit vector, $\hat{\mathbf{x}}_{kb}$, from branch $k$ to branch $b$ causing
them to move further apart; attraction causes the opposite movement. The
interaction, $\mathbf{I}_b$, acting on branch $b$, is given by
%
\begin{equation}
\mathbf{I}_b = \frac{1}{|B_b|} \sum_k \hat{\mathbf{x}}_{kb}\,Q(d_{kb})
\end{equation}
%
where $Q(d_{kb})$ is the signalling strength between two growth cones of
radius $r$ a distance $d_{kb}$ from one another, given by
%
\begin{equation}
Q(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r \\
     0 & d_{kb} > 2r
     \end{cases}
\end{equation}
%
The sign of $Q$ (which is dependent on the values of $F_i$) determines whether
the interaction, $\mathbf{I}_b$, is repulsive or attractive. $l_{i,k}$ is the
expression of ligand $i$ on branch $k$.

\subsection{Movement}

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, of each
branch, $b$, is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{I}_b
\end{equation}
%
evaluated at time $t$.

@@

>>>>

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum of
side length 1 in arbitrary units.
%
\citet{simpson_simple_2011} model three contributions to the movement of the growth
cones; chemoaffinity, competition for space and axon-axon interactions. The
first two of these are modelled without reference to a particular mechanism;
each agent is provided with globally-supervised information about its
instantaneous location and its target location, allowing it to move along a
vector towards the target. Competition is implemented as a movement of
constant distance along the vector between two agents if those two agents are
within a threshold distance of one another. Interaction is signal-induced
repulsion, which is triggered if two agents are both within the threshold
distance of each other and have compatible EphA receptor expression.

In our model, space-based competition was omitted and chemoaffinity and
axon-axon interactions were remodelled so that all interactions were via
signal transmission between receptors and ligands, which are expressed in
graded patterns on both retina and tectum. Signals induce attractive or
repulsive interactions; for chemoaffinity the interaction is to ascend or
descend the gradient of ligand expression on the tectum; for axon-axon
interactions, the strength of short-range repulsion (or attraction) is
governed by the receptor-ligand expression on pairs of axons.

\subsection*{Receptor and ligand expression}

@@

Each retinal ganglion cell projects $n=8$ growth cone agents (referred to
as \emph{branches}) which carry a set of 4 receptors, $r_i$, and 4 ligands,
$l_i$, indexed by $i$ and
expressed at levels determined by the cell's location on the retina. The
expression levels of the receptors and ligands vary with respect to the cell's
position on the retinal surface. Each receptor or ligand varies with respect
only to one dimension.  Receptor expression gradients are arranged in
orthogonal pairs with the gradient of $r_0$ being orthogonal to that of
$r_1$. $r_2$, whose gradient is opposite to $r_0$, is orthogonal to
$r_3$. Ligands are also arranged in orthogonal pairs, with $l_0$ opposing
$r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$ orthogonal to $l_0$ and
opposing $r_1$.

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression
as \citet{simpson_simple_2011} which is
\begin{equation}
r_i(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}

The simulated tectum expresses the same types of receptors and ligands, also
in orthogonal pairs of gradients. In this work, we did not consider the effect
of reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression, $L_i$. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
\textbf{X}). Correspondingly, we leave open the possibility that tectal ligand
expression may be modelled by exponential, linear or logarithmic functions:
\begin{equation}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}
or
\begin{equation}
L_i^{\text{lin}}(x,y) = 1.31 + 2.33 x;
\end{equation}
or
\begin{equation}
L_i^{\text{log}}(x,y) = 2.32 + 1.29 \log (2.3(x+0.2))
\end{equation}
(each of which has approximately the same expression values at $x=0$ and $x=1$).

@@

\subsection*{Movement}

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, of each
branch, $b$, is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect, $\mathbf{G}_b$ and an axon-axon interaction, $\mathbf{I}_b$:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{I}_b
\end{equation}
%
evaluated at time $t$. $m_1$ and $m_2$ are scalar parameters.

\subsection*{Chemoaffinity}

@@

\subsection*{Axon-axon interactions}

For axon-axon interactions, repulsion causes a movement of branch $b$ along a
unit vector, $\hat{\mathbf{x}}_{kb}$, from branch $k$ to branch $b$ causing
them to move further apart; attraction causes the opposite movement. The
interaction, $\mathbf{I}_b$, acting on branch $b$, is given by
%
\begin{equation}
% No \frac{1}{|B_b|} in this one
\mathbf{I}_b = \sum_k \hat{\mathbf{x}}_{kb}\,Q(d_{kb})
\end{equation}
%
where $Q(d_{kb})$ is the signalling strength between two growth cones of
radius $r$ a distance $d_{kb}$ from one another, given by
%
\begin{equation}
Q(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r \\
     0 & d_{kb} > 2r
     \end{cases}
\end{equation}
%
The sign of $Q$ (which is dependent on the values of $F_i$) determines whether
the interaction, $\mathbf{I}_b$, is repulsive or attractive. $l_{i,k}$ is the
expression of ligand $i$ on branch $k$.

\subsection*{Initial conditions}

Branches were randomly distributed in a stripe at the rostral side of the
tectum at the start of each simulation. Each RGC axon was assigned a random
initial position coordinate:
\begin{equation}
\mathbf{x}_{\text{axon},t=0} = (U(0,1), U(-0.2,0))
\end{equation}
where U(p,q) is a number selected from a random uniform distribution in the
range $[p,q)$. Each of the $n$ branches per RGC axon was given its parent
axon's initial position, plus a randomly generated offset with coordinates
derived from a normal distribution of mean 0 and standard deviation 0.1.

@@

\subsection*{Chemoaffinity model}

First, we describe the behaviour of the model with $m_2 = 0$, which has only
the gradient-based chemoaffinity effect and no axon-axon interaction.

\subsubsection*{Wildtype}
\subsubsection*{Tectal ablation etc}

\subsection*{Chemoaffinitiy with axon-axon interaction}

\subsubsection*{Wildtype}
\subsubsection*{Tectal ablation etc}
